CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-05-30 script/en/chapters/numbertheory.tex
fixes for beta6

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823$-$Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

% --------------------------------------------------------------------------
\subsubsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsubsection*{Example of an multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[h]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|ccccc}
Multiplication table modulo $6$: \quad* &  1 & 2 & 3 & 4 & 5  \\
\hline 
1 &  1 & 2 & 3 & 4 & 5 \\  
2 &  2 & {\bf 4} & {\bf 0} & {\bf 2} & 4 \\
3 &  3 & {\bf 0} & {\bf 3} & {\bf 0} & 3 \\
4 &  4 & {\bf 2} & {\bf 0} & {\bf 4} & 2 \\
5 &  5 & 4 & 3 & 2 & 1 \\
\end{tabular}  
\end{center} 
\end{table}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then takes the product $a*i {\rm ~mod ~} m $ for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}{chapter 3.9 Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

The inverses of the powers are also defined. The roots and logarithms are again whole numbers. Yet in contrast 
to the usual situation, they are not only difficult to calculate but, in the case of large numbers, cannot be 
calculated at all within a reasonable space of time.

@@

\begin{itemize}
\item The set is closed \index{Closeness}
\item A neutral element exists
\item An inverse element exists
\item The associative law applies.
\end{itemize}

@@

\begin{definition}\label{def-zth-phiofn} \hypertarget{EulerFunction}{} 
The Euler function\footnote{Is often described as the Euler phi function\index{Euler!(phi) function} $\Phi(n)$.} $J(n)$ specifies the number of 
elements in $\mathbb{Z}_n^*$.
\end{definition}
$J(n)$ also specifies how many whole numbers have multiplicative inverses in mod $n$.
$J(n)$ can be calculated if we know the prime factors of $n$\index{Prime factor!decomposition}.

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and movies\index{Movies} not only 
classical methods are used (like the secret message solved by Sherlock Holmes
in ``The Adventure of the Dancing Men'' by Arthur Conan  
Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999 (included in the CrypTool package
  \index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\item the novel ``Das Museum der gestohlenen Erinnerungen (Museum of
  the stolen memories'' by Ralf Isau \index{Isau 2003}, 
  Thienemann 2003, where the last part of the oracle can only be
  solved with the joined help of the computer community.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}
}
procedure in the group $\mathbb{Z}_n^*$.

@@

>>>>

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823 $-$ Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

% --------------------------------------------------------------------------
\subsubsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsubsection*{Example of an multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Multiplication table modulo $6$: \quad* &  1 & 2 & 3 & 4 & 5  \\
\hline 
1 &  1 & 2 & 3 & 4 & 5 \\  
2 &  2 & {\bf 4} & {\bf 0} & {\bf 2} & 4 \\
3 &  3 & {\bf 0} & {\bf 3} & {\bf 0} & 3 \\
4 &  4 & {\bf 2} & {\bf 0} & {\bf 4} & 2 \\
5 &  5 & 4 & 3 & 2 & 1 \\
\end{tabular}  
\end{center} 
\end{table}

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}{chapter 3.9 Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

The inverses of the powers are also defined. The roots and logarithms are again whole numbers. Yet in contrast 
to the usual situation, they are not only difficult to calculate but, in the case of large numbers, cannot be 
calculated at all within a reasonable amount of time.

@@

\begin{itemize}
\item The set is closed \index{Closeness}
\item A neutral element exists
\item An inverse element exists for each element
\item The associative law applies.
\end{itemize}

@@

\begin{definition}\label{def-zth-phiofn} \hypertarget{EulerFunction}{} 
The Euler function\footnote{Often written as the Euler phi function\index{Euler!(phi) function} $\Phi(n)$.} $J(n)$ specifies the number of 
elements in $\mathbb{Z}_n^*$.
\end{definition}
$J(n)$ also specifies how many whole numbers have multiplicative inverses in mod $n$.
$J(n)$ can be calculated if we know the prime factors of $n$\index{Prime factor!decomposition}.

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and movies\index{Movies} not only 
classical methods are used (like the secret message solved by Sherlock Holmes
in ``The Adventure of the Dancing Men'' by Arthur Conan  
Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999 (included in the CrypTool package
  \index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\item the novel ``Das Museum der gestohlenen Erinnerungen (Museum of
  the stolen memories)'' by Ralf Isau \index{Isau 2003}, 
  Thienemann 2003, where the last part of the oracle can only be
  solved with the joined help of the computer community.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}
}
procedure in the group $\mathbb{Z}_n^*$.

@@

The 160-digit decimal number origins from the old challenge list of RSADSI.
This number was retracted after RSA-155 (RSA512) had been factorized successfully,
leaving RSA576, RSA640 etc. (using 64 bit steps upwards).
The prime factors of RSA-160 were still unknown.
So this record of the team of Prof.\ Franke provides the solution of 
the old challenge, for which RSADSI didn't award a price anymore.
